% ============================== BIT ALGORITHMS ============================== %
% Project:          The Experimental Bit Algorithms Library
% Name:             bit-algorithms.tex
% Description:      Scientific and technical paper on bit algorithms
% Creator:          Vincent Reverdy
% Contributor(s):   Vincent Reverdy [2019]
% License:          BSD 3-Clause License
% ============================================================================ %



% =============================== CONFIGURATION ============================== %
% Document class
\documentclass[letterpaper, 8pt, twocolumn]{article}
% ---------------------------------------------------------------------------- %
% Packages
\usepackage[
    tmargin = 1in,
    bmargin = 1in,
    lmargin = 1in,
    rmargin = 1in
]{geometry}
\usepackage{lipsum}
\usepackage{lmodern}
\usepackage{titling}
\usepackage{titlesec}
\usepackage[sf=false]{libertine}
\usepackage{cite} \usepackage{array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{float}
\usepackage{fdsymbol}
\usepackage{tabularx}
\usepackage{epigraph}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{siunitx}
\usepackage{colortbl}
\usepackage{enumitem}
\usepackage[listings,skins,theorems,breakable,most]{tcolorbox}
\edef\restoreparindent{\parindent=\the\parindent\relax}
\usepackage[parfill]{parskip}
\restoreparindent
\usepackage{hyperref}
% ---------------------------------------------------------------------------- %
% Compilation
\pdfpageattr{/Group << /S /Transparency /I true /CS /DeviceRGB>>}
% ---------------------------------------------------------------------------- %
% Bibliography
\bibliographystyle{apalike}
% ============================================================================ %



% ================================== LAYOUT ================================== %
% Page style
\pagestyle{fancy}
\fancyhead{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
% ---------------------------------------------------------------------------- %
% Fonts
\renewcommand{\ttdefault}{cmtt}
% ---------------------------------------------------------------------------- %
% Spacing
\titlespacing\section{0pt}{0.25\baselineskip plus 2pt minus 1pt}{0.125\baselineskip plus 2pt minus 1pt}
\titlespacing\subsection{0pt}{0.25\baselineskip plus 2pt minus 1pt}{0.125\baselineskip plus 2pt minus 1pt}
\titlespacing\subsubsection{0pt}{0.25\baselineskip plus 2pt minus 1pt}{0.125\baselineskip plus 2pt minus 1pt}
\setlist{topsep=-0.25\baselineskip, itemsep=0.25\baselineskip, partopsep=0.\baselineskip, parsep=0.\baselineskip}
% ============================================================================ %



% ================================= COMMANDS ================================= %
% Acronyms
\newcommand{\acronym}[1]{\textsc{#1}}
\newcommand{\acrolang}[1]{\textsc{#1}}
\newcommand{\cpp}{\acrolang{C++}}
\newcommand{\clang}{\acrolang{C}}
\newcommand{\typecell}[3]{\cellcolor{#1}\textcolor{blue}{#2}#3}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
% ============================================================================ %



% ================================ INFORMATION =============================== %
% Information
\title{Towards a Standard Library of Bit Manipulation Algorithms}
\author{%
Vincent Reverdy%
%\and FirstName LastName%
}
\date{\today}
% ---------------------------------------------------------------------------- %
% Hyperref
\hypersetup{
    pdfauthor={\theauthor},
    pdftitle={\thetitle},
    pdfsubject={Bit manipulation algorithms},
    pdfkeywords={high performance computing, programming languages, algorithms, bit, bit manipulation, C++},
    pdfproducer={LaTeX},
    pdfcreator={pdflatex},
    colorlinks,
    citecolor=blue,
    filecolor=blue,
    linkcolor=blue,
    urlcolor=blue
}
% ============================================================================ %



% =================================== MAIN =================================== %
\begin{document}
\maketitle
% ---------------------------------------------------------------------------- %
\begin{abstract}
\label{sec:abstract}
\end{abstract}
% ---------------------------------------------------------------------------- %
\section{Introduction}
\label{sec:introduction}
% ---------------------------------------------------------------------------- %
\section{Algorithms}
\label{sec:algorithms}

\paragraph{Notation}


\subsection{Non-modifying Algorithms}
\label{subsec:nonmodifyingAlgs}
\subsubsection{\texttt{max\_element}}
\paragraph{Description}
Locates the largest element in a range \texttt{[first, last)} and returns an iterator to the element.

\paragraph{Implementation}
Implementation is somewhat trivial. If \texttt{first} is LSB aligned (meaning it points to the 0th bit of a word) we perform a partial read from \text{first} to the closer of the MSB and the \text{last} iterator. Missing bits from the read are padded with 0 bits in the versions of the algorithm where a custom comparator is not passed. Next, we check the read word to see if it is greater than zero. If it is greater than zero, we know there is a set bit somewhere within the word. We can find the exact position of the set bit using \texttt{\_tzcnt} and return this position as an iterator. If \texttt{first} is not equal to \texttt{last}, we advance first to the 0th bit of the next word.

While \texttt{first} is not equal to \texttt{last} and \texttt{last} is not in the same word as \texttt{first} (marking a final and partial word), we read whole words as normal and check for set bits as previously described.

If we arrive at a final, partial word, we again take a partial read of the bits remaining in the range, padding any missing bits with 0. We check for set bits as normal and return an iterator if a set bit was found.

If we have scanned the entire range and no set bits were found, we simply return \texttt{first}.

\subsubsection{\texttt{replace}}
\paragraph{Description}
Replaces 0s with 1s or vice versa in a range \texttt{[first, last)}.

\paragraph{Implementation}
Implementation follows a typical sequential processing of words in the range. First, the word pointed to by \texttt{first} is processed. Next, any full intermediary words within \texttt{[first, last)} are processed. Finally, if there is a remaining partial word bounded by \texttt{last}, it is handled.

For full words, replacement is trivial. If we are replacing 0s with 1s, we simply set each word equal to -1. If we are replacing 1s with 0s, we set each word equal to 0. Partial words are slightly more complicated. For some processed partial word starting at bit index $i$ and ending at index $j$, we create a mask where bits are unset between $i$ and $j$ and set outside of the range. If we are replacing 0s with 1s, we do a bitwise or between the full word being processed and the mask. If we are replacing 1s with 0s, we do a bitwise and between the full word being processed and the XOR of the mask. This replaces bits within the bounds of the range while not modifying any bits outside the range.

\subsubsection{\texttt{min\_element}}
\paragraph{Description}
Locates the smallest element in a range \texttt{[first, last)} and returns an iterator to the element.

\paragraph{Implementation}
Implementation follows a typical sequential processing of words in the range. First, the word pointed to by \texttt{first} is processed. Next, any full intermediary words within \texttt{[first, last)} are processed. Finally, if there is a remaining partial word bounded by \texttt{last}, it is handled.

To locate 0's, we read words and pad any missing bits with 1s. We use a bitwise not to flip all of the bits so that we can use \texttt{\_tzcnt} (which counts the number of 0s from LSB to the first 1) to find the position of the first 0. We return an iterator to this position or \texttt{first} if no zero exists.


\subsection{Modifying Algorithms}
\label{subsec:modifyingAlgs}

\subsubsection{\texttt{copy}}
\label{subsubsec:copy}
\paragraph{Description}
Copies all elements in the range [first, last) starting from first and 
proceeding to last - 1. The behavior is undefined if d\_first is within the 
range [first, last). In this case, std::copy\_backward may be used instead.

\paragraph{Implementation}
See psuedocode below:
\begin{algorithm}[H]
    \caption{Current copy implementation}
    \begin{algorithmic}[1]
        \Function{copy}{\texttt{first1, last1, first2}}
        \If {!\texttt{is\_aligned}(\texttt{first2})}
            \State Use \texttt{get\_word} to copy bits and align \texttt{first2}
            \State Advance \texttt{first1} correspondingly.
        \EndIf
        \While {There are remaining bits to copy} 
            \State Construct words of \texttt{dst\_word\_type}
            from the source and assign to destination.
        \EndWhile
        \State If the last iterator is not aligned, copy the remaining bits to the
            final word.
        \EndFunction
    \end{algorithmic}
\end{algorithm}
\paragraph{Comments}
\begin{itemize}
    \item Might be worth thinking about having separate implementations 
        depending on the size of the words using \texttt{if constexpr}.
\end{itemize}

\subsubsection{\texttt{shift\_left} and \texttt{shift\_right}}
\label{subsubsec:shift}
\paragraph{Description}
\begin{itemize}
    \item[\texttt{shift\_left}] Shifts the elements towards the beginning of the 
            range. If $n \leq 0 || n \geq last - first$, there are no effects. 
            Otherwise, for every integer $i$ in $[0, last - first - n)$, moves the 
            element originally at position $first + n + i$ to position $first + i$. 
            The moves are performed in increasing order of i starting from.
    \item[\texttt{shift\_right}] Shifts the elements towards the end of the 
        range. If $n \leq 0 || n \geq last - first$, there are no effects. Otherwise, 
        for every integer $i$ in $[0, last - first - n)$, moves the element 
        originally at position $first + i$ to position $first + n + i$. If ForwardIt 
        meets the LegacyBidirectionalIterator requirements, then the moves are 
        performed in decreasing order of $i$ starting from $last - first - n - 1$.
\end{itemize}

\paragraph{Implementation}
For both left and right shift, let $N$ be the number of total word shifts and 
$n'$ be the remainder of needed bit shifts i.e.
$$N=\floor*{\frac{n}{|\texttt{word\_type}|}}$$ $$n'=n-N*|\texttt{word\_type}|$$
\subparagraph{\texttt{shift\_left}}

\begin{algorithm}[H]
    \caption{Current \texttt{shift\_left} implementation}
    \begin{algorithmic}[1]
        \Function {shift\_left}{\texttt{first, last, n}}
            \State Save first and last words if not aligned.
            \State Unset bits in first word that are not part of 
            the range.
            \State Use STL version to shift words by $N$.
            \State Create latent iterator to keep track of penultimate word.
            \State \texttt{\_shrd} each word and the next by $n'$ until we 
            reach the last word
            \State Right shift the last word by $n'$.
            \State Put back the out of range bits if necessary.
        \EndFunction
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{Current \texttt{shift\_right} implementation}
    \begin{algorithmic}[1]
        \Function {shift\_left}{\texttt{first, last, n}}
            \State Save first and last words if not aligned.
            \State Unset bits in last word that are not part of 
            the range.
            \State Use STL version to shift words by $N$.
            \State Create two temporary variables \texttt{T1} and \texttt{T2}
            \State \texttt{T1 = *first.base()}
            \State \texttt{*first.base() <<= n'}
            \For {\texttt{it=first.base(), it <= last, it++}}
                \State \texttt{T2 = *it}
                \State \texttt{*it = \_shld(*it, T1, n')}
                \State \texttt{T1 = T2}
            \EndFor
            \State Put back the out of range bits if necessary.
        \EndFunction
    \end{algorithmic}
\end{algorithm}

\paragraph{Comments}
\begin{itemize}
    \item \texttt{shift\_right} even with a ForwardIt may be able to avoid
        having \textbf{two} temporary variables.
    \item If we have a bidiriectional iterator, \texttt{shift\_right} 
        can avoid temporary variables and just use \texttt{\_shld} starting 
        from last and going to first.
\end{itemize}
% ---------------------------------------------------------------------------- %
\bibliography{bit-algorithms}
% ---------------------------------------------------------------------------- %
\end{document}
% ============================================================================ %



